Kindly Bent to Free Us

### 2019-07-10 kindly/lang.tex
type environments

\subsubsection{Environments and bindings}
\label{sdtyping:envs}

\lang aims to control the usage of variables, either
through
linearity and affinity which restrict how variables are shared and ignored,
or through borrows which allow to circumvent the linearity rules
in a controlled way.
In order to check these properties, we introduce several new bindings
in our typing environments.
The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.
On top of the traditional type and value bindings, noted respectively
$\bvar{\tvar}{\kschm}$ and $\bvar{x}{\schm}$, we also consider two new
bindings.
Borrow bindings $\bvar{\borrow{x}}{\borrowty{k}{\tau}}$ indicates
that the borrow $\borrow{x}$ can be used directly, as implemented
by the rule {\sc Borrow} in \cref{selectrules:borrow}.
Suspended bindings, noted $\svar{x}{\schm}^n$, indicates that $x$
cannot be used immediately, but will be usable in a region
marked with $x$, which we detail in \cref{sdtyping:regions}.

Environments are also used to encode several linearity-related properties
through \emph{environment-wide constraints}.
Here we present how to control both unused variables and
closure captures through such environment constraints, noted $\Cleq{\E}{k}$.
The intuitive meaning of $\Cleq{\E}{k}$ is that for all bindings
$\bvar{x}{\tau}$ in $\E$,
if type $\tau$ has kind $k'$ (noted $\inferSK{C}{\E}{\tau}{k'}$)
then the following constraints should hold: $\Cleq{k'}{k}$.
Borrow bindings follow the exact same rules, but suspended bindings
are forbidden in such constraints.
While this definition is incomplete, it is sufficient to understand
its use in the {\sc Abs} and {\sc Var} rules shown in \cref{selectrules:binders}.
%
The {\sc Var} rule looks up the type scheme of the variable $x$ in
the environment $\E$
and instantiate a concrete type through the substitution $\unif$. It then
checks that the other bindings in $\E$ can safely be discarded by
ensuring they are not linear. This is achieved through the constraint
$\Cleq{\E\Sdel{x}}{\kaff_\infty}$ which replaces the traditional
``weakening'' rule found in many type systems.

The {\sc Abs} rule must ensures that if the body of the lambda captures
linear or affine variables, the content will be reflected in the kind
annotated on the arrow, such as $\tau_2 \tarr{k} \tau_1$.
In order to do so, we constraint all the bindings of $\E$ to be less than $k$.
If one of the binding of $\E$ contains a type whose kind is affine,
we will derive the constraints $\Cleq{\kaff}{k}$, which ensures that
the kind on the arrow is at least affine.
Capturing a borrow is perfectly allowed here: the kind of the borrow
will be reflected on the arrow which will restrict the closure
to the region of the borrow. Capturing a \emph{suspended} binding, however,
is forbidden.
Note that $x$ is \emph{not} included in $\E$: since it's the argument of the
closure, its use should not induce a capture.

\begin{figure}[!h]
  \begin{minipage}{0.33\linewidth}
    \begin{align*}
      \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
      B &::=\ \bnone \tag{Empty} \\
         &\mid \bvar{x}{\schm}\tag{Variables}\\
         &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
         &\mid \svar{x}{\schm}^n \tag{Suspended}\\
         &\mid \bvar{\tvar}{\kschm}\tag{Types}
         % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
    \end{align*}
    \caption{Type environments}
    \label{grammar:env}
  \end{minipage}\hfill
  \begin{minipage}{0.65\linewidth}
    \begin{mathpar}
      \ruleSDBorrow
      \and
      \ruleSDLam
      \and
      \ruleSDVar
    \end{mathpar}
    \caption{The {\sc Borrow}, {\sc Abs} and {\sc Var} rules}
    \label{selectrules:borrow}
    \label{selectrules:binders}
  \end{minipage}
\end{figure}

>>>>

\subsubsection{Environments and bindings}
\label{sdtyping:envs}
\begin{figure}[tp]
  \begin{minipage}{0.33\linewidth}
    \begin{align*}
      \E &::=\ \Eempty \mid \E;B\tag{Environments}\\
      B &::=\ \bnone \tag{Empty} \\
         &\mid \bvar{x}{\schm}\tag{Variables}\\
         &\mid \bvar{\borrow{x}}{\borrowty{k}{\tau}}\tag{Borrows}\\
         &\mid \svar{x}{\schm}^n \tag{Suspended}\\
         &\mid \bvar{\tvar}{\kschm}\tag{Types}
         % |&\ \tydecl{t}{\kschm}{K}{\tau}\tag{Type Declaration}\\
    \end{align*}
    \caption{Type environments}
    \label{grammar:env}
  \end{minipage}\hfill
  \begin{minipage}{0.65\linewidth}
    \begin{mathpar}
      \ruleSDBorrow
      \and
      \ruleSDLam
      \and
      \ruleSDVar
    \end{mathpar}
    \caption{The {\sc Borrow}, {\sc Abs} and {\sc Var} rules}
    \label{selectrules:borrow}
    \label{selectrules:binders}
  \end{minipage}
\end{figure}

\lang controls the use of variables by supporting new modes of
binding in a type environment $\E$.
Its definition is given in  \cref{grammar:env}.
In addition to standard bindings of type variables to kind schemes,
$\bvar{\tvar}{\kschm}$,  and value bindings, $\bvar{x}{\schm}$, there
are suspended bindings and borrow bindings.
A suspended binding, $\svar{x}{\schm}^n$, indicates that $x$ is
earmarked for a borrowing use in a nested region
marked with $x$ (see \cref{sdtyping:regions}), but
cannot be used directly.
A borrow binding, $\bvar{\borrow{x}}{\borrowty{k}{\schm}}$, replaces
such a suspended binding on entry to the $x$-region. It indicates 
that the borrow $\borrow{x}$ can be used directly. The kind $k$
restricts the lifetime of the borrow to the region (see \cref{selectrules:borrow}).

% , either
% through
% linearity and affinity which restrict how variables are shared and ignored,
% or through borrows which allow to circumvent the linearity rules
% in a controlled way.
% In order to check these properties, we introduce several new bindings
% in our typing environments.
% The grammar of type environment (noted $\E$) is shown in \cref{grammar:env}.

% Environments are also used to encode several linearity-related properties
% through \emph{environment-wide constraints}.
Constraints on an environment control substructural properties by
restricting the types of variables.  The constraint $\Cleq{\E}{k}$
abbreviates $\Cleq\schm k$, for all $\bvar{x}{\schm}$
% $\svar{x}\schm$, and $\bvar{\borrow{x}}{\borrowty{k'}{\schm}}$
in $\E$, which in turn means that $k' \lk k$ where $k'$ is the kind of
the variable's type scheme.
Borrow bindings follow the exact same rules, but suspended bindings
are forbidden in an environment $\E$ constrained like that.
This intuitive explanation is sufficient to understand
the {\sc Abs} and {\sc Var} rules shown in
\cref{selectrules:binders}.

The {\sc Var} rule looks up the type scheme of the variable $x$ in
the environment $\E$
and instantiates it using the substitution $\unif$. The rule also
checks that the other bindings in $\E$ can be safely discarded by
imposing the constraint $\Cleq{\E\Sdel{x}}{\kaff_\infty}$. 
It enforces that all remaining bindings (except $x$) are affine or
unrestricted and can therefore be discarded.

The {\sc Abs} rule ensures that the kind annotation on the arrow type
($\tau_2 \tarr{k} \tau_1$) reflects the restrictions on the captured variables.
To do so, the rule imposes the constraint $\Cleq\E k$.
If, for instance, any binding in $\E$ is affine, it gives rise to the
constraint $\Cleq{\kaff_n}{k}$, so that  the arrow kind is at least
affine at nesting level $n$.
Capturing a borrow is perfectly fine: the kind of the borrow is also a
lower bound of the arrow kind $k$ which restricts the closure
to the region of the borrow.
Capturing a suspended binding is forbidden.
% Note that $x$ is \emph{not} included in $\E$: since it's the argument of the
% closure, its use should not induce a capture.
